% conclusions.tex — §8 (merged discussion + conclusions)

\section{Discussion and Conclusions}
\label{sec:conclusions}

We have applied Singular Value Decomposition to DESI DR2 BAO measurements and three Type~Ia supernova compilations (Union3, Pantheon+, DES-Dovekie), all analyzed with a common CMB baseline from ACT DR6. Our main findings are:

\begin{enumerate}

\item \textbf{Universal \czero.} All low-$z$ distance probes measure essentially the same leading SVD direction $V_0$ (cross-probe inner products $> 0.996$); we call its data amplitude \czero. Within \LCDM, this single number captures the leading low-$z$ distance information from each probe --- \czero is, to high accuracy, a measurement of \omh.

\item \textbf{The \wowa preference is primarily \czero.} The DESI \wowa preference is dominated by \czero; the BAO dark energy direction (\cone) shows only $+1.2\sigma$ tension with \LCDM. At the BAO \cone pivot ($z = 0.46$), $w = -0.94 \pm 0.052$ ($+1.2\sigma$)---consistent with a cosmological constant. SN \cone tensions are also mild ($-1.5\sigma$ to $+0.2\sigma$). BAO prefers lower \omh than the CMB ($+2.2\sigma$). The three SN datasets (DES-Dovekie $-0.8\sigma$, Pantheon+ $-1.1\sigma$, Union3 $-1.7\sigma$) do not have enough constraining power in \czero to either confirm or refute the BAO result.

\item \textbf{Curvature coherence.} Spatial curvature is the only other extension that opens a new BAO direction, and only marginally. Both BAO measurable directions (\czero and \cone) independently prefer positive spatial curvature --- an interesting internal consistency check --- although the \cone error bar is very large. SN data give $\sigres < 0.1$ for $\Omk$ and so cannot provide an independent cross-check.

\item \textbf{No other extension is distinguishable.} No other tested extension (EDE, $A_\mathrm{lens}$, $B_\mathrm{PMF}$) opens a measurable new direction ($\sigres < 0.3$). Low-$z$ distance data cannot distinguish among them.
 
\item \textbf{CMB lensing contribution and $\tau$.} The CMB \omh constraint has a lensing component: freeing $A_\mathrm{lens}$ both broadens and shifts the \omh posterior, reducing the BAO \czero tension from $2.2\sigma$ to $1.1\sigma$. This is the same physics as shifting the optical depth $\tau$, which \citet{Sailer2026} have shown the rest of the CMB + BAO data prefer to be $\sim 3$--$5\sigma$ higher than the Planck low-$\ell$ value ($\tau = 0.090 \pm 0.012$ versus $0.054 \pm 0.007$). At the current level of precision, a systematic in Planck's $\tau$ would be a single-point failure for many cosmological analyses.

\item \textbf{The lowest-redshift BAO point.} A striking feature of the BAO data is the high $\DM/\DHubble$ value of the lowest-redshift anisotropic point. Fitting it requires the higher SVD modes ($c_2$, $c_3$) beyond the leading dark-energy direction (Figure~\ref{fig:bao_data_bands}, right panel), which within a dark-energy interpretation translates into a rapid evolution of $w(z)$. This makes that single point unusually influential. Any improvement in the next DESI data release---a smaller error bar on this point, or a measurement bringing comparable-precision information from lower redshift---would therefore be especially valuable for establishing whether this feature is real.

\end{enumerate}

The SVD framework tells us \emph{where} the tension is (\czero, i.e., \omh) but not \emph{what} causes it. This is a fundamental limitation of distance measurements, not a shortcoming of the method: the SVD transparently identifies the boundary between what low-$z$ probes can and cannot constrain.

All current low-redshift distance data, whether from BAO or supernovae, are effectively one-dimensional in their cosmological information content. That single dimension is well described by \omh.
